---
title: "Leaflet_football_clubs"
author: "Simon Larsen and Jesper Jensen"
date: "2026-03-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Introduction
In our project, we have chosen to create a map showing the locations of the 25 largest football clubs in Denmark. Our selection is based on which league the respective clubs will play in during the 2025/26 season. On this basis, we have included the 24 clubs competing in Denmark’s two top leagues—the Superliga and the 1st Division—plus Skive IK from the 2nd Division, Denmark’s third-highest tier. We included Skive IK because it is our childhood club, and adding it also brings the total to a round number.
In our spreadsheet, we have listed the following information about the 25 clubs: the club’s name, the name of its stadium, and the league in which the club competes. For the club’s specific location, we used the location of its stadium and recorded the corresponding longitude and latitude coordinates.



########################################  TASK NUMBER ONE


# Task 1: Create a Danish equivalent of AUSmap with Esri layers, 
# but call it DANmap. You will need it layer as a background for Danish data points.
```{r}

library(tidyverse)
library(googlesheets4)
library(leaflet)

esri <- grep("^Esri", providers, value = TRUE)


DANmap<- leaflet() %>%   # Creates a new empty Leaflet map object
  setView( 10.85089,55.2339084, zoom = 6.25) %>% # Sets the initial map center and zoom level over Denmark
  addTiles() # Adds the defaultOpenStreetMap background layer




 DANmap1<- DANmap %>%
  addLayersControl(baseGroups = names(esri), # Adds a menu that lets the user switch between Esri basemap layers
                   options = layersControlOptions(collapsed = FALSE)) %>% # Keeps the layer menu expanded by default
  addMiniMap(tiles = esri[[1]], toggleDisplay = TRUE,# Displays a small minimap in the bottom‑right corner
             position = "bottomright") %>%
  addMeasure(
    position = "bottomleft", # Places the measurement tool in the bottom‑left corner
    primaryLengthUnit = "meters",# Measures distances in meters
    primaryAreaUnit = "sqmeters",# Measures areas in square meters
    activeColor = "#3D535D",# Color used while a measurement is active
    completedColor = "#7D4479") %>% # Color used when a measurement is completed
  htmlwidgets::onRender(" # Adds JavaScript that syncs the minimap with the selected basemap
                        function(el, x) {
                        var myMap = this;
                        myMap.on('baselayerchange',
                        function (e) {
                        myMap.minimap.changeLayer(L.tileLayer.provider(e.name));
                        })
                        }") %>% 
  addControl("", position = "topright") # Adds an empty control box (used for layout spacing)


 DANmap1
``` 
 



######################################## TASK TWO


# Task 2: Read in the googlesheet data you and your colleagues created
# into your DANmap object (with multiple background layers you created in Task 1).

# Solution
```{r}
# Libraries
install.packages("googlesheets4")
library(tidyverse)  # Loads tidyverse packages for data handling and analysis
library(googlesheets4) # Enables reading data directly from Google Sheets
library(leaflet) # Loads Leaflet for creating interactive maps



football_stadiums <- read_sheet( # Reads a Google Sheet into an R dataframe
  "https://docs.google.com/spreadsheets/d/1EcYYLqle10mE7E8Rbw1HfFMo0y7HX5HPJjb6hwNl0nw",
  sheet = "Sheet1", # Specifies which sheet to import
  col_types = "cccnnc" # Defines column types: 3 character, 2 numeric, 1 character
)



glimpse(football_stadiums) 


football_stadiums %>% # Starts a pipeline using the dataset
  filter(!is.na(Longitude)) %>% # Removes rows where longitude is missing
  filter(!is.na(Latitude)) # Removes rows where latitude is missing


football_stadiums <- football_stadiums |> # Continues working with the dataset
  dplyr::rename(  # Renames columns to temporarily store incorrect coordinates
    WrongLat = Latitude, # Saves incorrect latitude values
    WrongLon = Longitude # Saves incorrect longitude values
  ) |> 
  dplyr::mutate( # Creates corrected coordinate columns
    Latitude = WrongLon, # Swaps longitude into latitude
    Longitude = WrongLat # Swaps latitude into longitude
  )



leaflet(football_stadiums) %>% # Initializes a Leaflet map using the dataset
  addTiles() %>% # Adds the default OpenStreetMap background
  addMarkers( # Adds markers for each football club
    lng = ~Longitude, # Uses longitude for marker placement
    lat = ~Latitude, # Uses latitude for marker placement
    popup = ~paste0(  # Defines popup text shown when clicking a marker
      "<b>", Club, "</b><br>",  # Displays club name in bold
      "Stadium: ", Stadium, "<br>", # Shows stadium name
      "League: ", League # Shows league name
    )
  )
```

######################################## TASK THREE

# Task 3: Can you cluster the points in Leaflet?
# Hint: Google "clustering options in Leaflet in R"

# Solution
```{r}
leaflet(football_stadiums) %>% # Creates a Leaflet map with clustering
  addTiles() %>%  # Adds the standard background map
  addMarkers( # Adds markers for all clubs
    lng = ~Longitude, # Longitude for marker placement
    lat = ~Latitude, # Latitude for marker placement
    popup = ~paste0( # Popup text for each marker
      "<b>", Club, "</b><br>", # Club name in bold
      "Stadium: ", Stadium, "<br>", # Stadium name
      "League: ", League # League name
    ),
    clusterOptions = markerClusterOptions() # Enables clustering to group nearby markers
  )
  
```
######################################## TASK FOUR

# Task 4: Look at the two maps (with and without clustering) and consider what
# each is good for and what not.
"The map with clustering is more pleasing to the eye,
because it reduces visual clutter and prevents markers from overlapping in areas
where many football clubs are located close together.
This creates a cleaner and more readable national overview,
especially in regions like Copenhagen or Aarhus where several clubs are concentrated.
However, clustering also requires the user to interact with the map
by zooming in to reveal the precise locations of individual stadiums.
In general, clustering is a strong choice when the goal is to present a large dataset
without overwhelming the viewer. It highlights regional patterns and makes the map
look more structured and professional. On the downside, clustering can hide
important details until the user zooms in, which means that exact stadium positions
are not immediately visible. For analytical tasks where precision matters,
a non‑clustered map may therefore be more appropriate, while clustering excels
when clarity and visual simplicity are the main priorities."



######################################## TASK FIVE

# Task 5: Find out how to display the notes and classifications column in the map. 
# Hint: Check online help in sites such as 
# https://r-charts.com/spatial/interactive-maps-leaflet/#popup

# Solution
```{r}
leaflet(football_stadiums) %>%  # Creates a Leaflet map including notes
  addTiles() %>% # Adds the default map tiles
  addMarkers( # Adds markers for each club
    lng = ~Longitude, # Longitude for marker placement
    lat = ~Latitude, # Latitude for marker placement
    popup = ~paste0( # Popup text including additional information
      "<b>", Club, "</b><br>", # Club name in bold
      "Stadium: ", Stadium, "<br>", # Stadium name
      "League: ", League, "<br>", # League name
      "Notes: ", Notes, "<br>" # Adds the notes column to the popup
    ),
    clusterOptions = markerClusterOptions() # Enables clustering for cleaner visualization
  )

```
######################################## CONGRATULATIONS - YOU ARE DONE :)
